Spitzer, Gaia and the Potential of the Milky Way

### 2013-07-19 text/paper1/paper.tex

\section{Context and motivation}
The method presented in Section \ref{sec:method} takes advantage of
three distinct developments in the observations of tidal streams: (i)
the demonstration of a technique for deriving distances to individual
RR Lyrae stars with 2\% accuracies (Section \ref{sec:spitzer}); (ii)
the prospect of proper motion measurements of the same stars with
$\sim$10~mas/yr precision (Section \ref{sec:gaia}); and (iii) the
tracing of debris associated with Sgr around the entire Galaxy
(Section \ref{sec:sgr})

@@

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia2001}, planned for launch in October
2013, is an astrometric mission which aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
over the 5-year lifetime of the mission, Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full six-dimensional phase-space
maps of the Galaxy with $<$10\% distance errors for heliocentric distances of up to $\sim$10~kpc for ...

Figure~\ref{fig:gaia_errors} show the Gaia end-of-mission distance and
tangential velocity error estimates for a typical halo RR Lyrae star
\citep[{$M_V=0.45$, [Fe/H]$=-1.5$};][]{ivezic2008}. Within 2~kpc, Gaia
will be able to measure distances to these stars with better than 2\%
accuracy: RR Lyraes in this volume will be used to test and calibrate
the Spitzer PL relation described above. Beyond the 2~kpc threshold,
the mid-IR PL relation for RR Lyraes will provide more accurate
distance measurements.

The anticipated proper motion errors for RR Lyrae correspond to
tangential velocity errors of order $\sim$1~km/s at 30~kpc, and of
order $\sim$10~km/s out to several tens of kiloparsecs. The
combination of Spitzer and Gaia data will extend the `horizon' of
where full six-dimensional phase-space maps of the Galaxy are possible
from $<10$~kpc to 80~kpc. This enormous increase in volume will refine
data on existing debris systems such as the Sagittarius stream.

>>>>

\section{Context and motivation}
The method presented in Section \ref{sec:method} takes advantage of
three distinct developments in the observations of tidal streams: (i)
the demonstration of a technique for deriving distances to individual
RR Lyrae stars with 2\% accuracies (Section \ref{sec:spitzer}); (ii)
the prospect of proper motion measurements of the same stars with
$\sim$10~$\mu$as/yr precision (Section \ref{sec:gaia}); and (iii) the
tracing of debris associated with Sgr around the entire Galaxy
(Section \ref{sec:sgr})

@@

\subsection{Gaia and the age of astrometry}
\label{sec:gaia}
The Gaia satellite \citep{gaia2001}, planned for launch in October
2013, is an astrometric mission which aims to measure the positions of billions of stars with
10-100~$\mu$as accuracies. With over 50 observations of the entire sky
over the 5-year lifetime of the mission, Gaia will
revolutionize the field of Galactic Astronomy by providing sufficient
proper motion accuracy to create full six-dimensional phase-space
maps of the Galaxy with $<$10\% distance errors for heliocentric distances up to $\sim$6~kpc for RR Lyrae stars.

Figure~\ref{fig:gaia_errors} show the Gaia end-of-mission distance and
tangential velocity error estimates for RR Lyrae. The velocity uncertainty depends on the magnitude of the source, which depends on the metallicity: each line in Figure~\ref{fig:gaia_errors} is computed by Monte Carlo sampling from the empirical metallicity distribution of the Galactic halo from \cite{ivezic2008} (centered around [Fe/H] $\sim$ -1.5). Within 2~kpc, Gaia
will be able to measure distances to these stars with better than 2\%
accuracy: RR Lyraes in this volume will be used to test and calibrate
the Spitzer PL relation described above. Beyond the 2~kpc threshold,
the mid-IR PL relation for RR Lyraes will provide more accurate
distance measurements.

\begin{figure}[h]
\begin{center}
\includegraphics[width=1.\textwidth, trim=2 2 5 5, clip]{/Users/adrian/projects/streams/plots/paper1/gaia.pdf}
\caption{ todo }\label{fig:gaia_errors}
\end{center}
\end{figure}

The anticipated proper motion errors for RR Lyrae correspond to
tangential velocity errors of order $\sim$1~km/s at 30~kpc, and of
order $\sim$10~km/s out to several tens of kiloparsecs. The
combination of Spitzer and Gaia data will extend the `horizon' of
where full six-dimensional phase-space maps of the Galaxy are possible
from $<10$~kpc to 80~kpc. This enormous increase in volume will greatly refine
data on existing debris systems such as the Sagittarius stream.
